---
title: "Commamox  analysis"
author: "Joshua Johnson"
output:
  html_document:
    df_print: paged
---

# Packages, data loading etc. 

```{r Packages, echo=TRUE, results='hide', warning=FALSE, message = FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(cowplot)
library(reshape2)
library(rstatix)
library(ggpubr)
library(nlme)
library(lme4)
```


```{r  qPCR load, echo=FALSE, results='hide', warning=FALSE, message = FALSE}

# Loading bay

  # qPCR data

   qpcr <- 
      read_delim("C:/Users/jjohn/OneDrive/MScthesis/data/comm.csv",",",)

  # DNA concentration

    dna <- 
      read_delim("C:/Users/jjohn/OneDrive/MScthesis/data/dna.csv",",",)
    
    trt <-
      read_delim("C:/Users/jjohn/OneDrive/MScthesis/data/treatment_factors.csv",",",) 
     
    
  # Variable type changes 
    
    qpcr$id <- as.factor(qpcr$id)
    dna$id <- as.factor(dna$id)
    
    trt <- trt %>%
      convert_as_factor(id,fert,fung,block,time,row)
    
```    
# Normalization on different standard curve (Comamox run specifics)

For commamox only one standard curve is viable. Therefore I will calculate quantity values based on the viable standard curve and correct the ct values of the other runs on the standard curve. 
```{r commamox: using one common standard curve calculations, echo=FALSE, result='hide'}
# Standard curve based on '[Type:Soil | Timepoint: T3] standard results

  standard_curve <- 
    read_delim("C:/Users/jjohn/OneDrive/MScthesis/data/qPCR/com_standard_curve.csv",",",)
  
  # Gettin an exacter model equation
  model <- lm(ct ~ qty_log10, data = standard_curve)
  model
  #Coefficients:
  #(Intercept)    qty_log10  
  #40.366       -3.648  


# Recalculating the quantity values based on the standard curve in '[Type:Soil | Timepoint: T3]

  # Load the data 
  
  std <- qpcr %>%
    filter(id == "Standard 2")%>%
    select(type,time,id,ct)%>%
    group_by(time,type)%>%
    get_summary_stats(ct, type = "mean_sd")
  
# We use the Standard 2 as an interrun calibrator and the Standard 2 of [Type:Soil | Timepoint: T3] is the 1.0 (100%)
  
  #Value is [Type:Soil | Timepoint: T3] 13.188
  
 std <- std %>%
    mutate(corr = mean - 13.188 )%>%
    select(time, type, corr)
 
 write_csv(std, "corr_factor.csv")
 
 qpcr <- 
   read_delim("C:/Users/jjohn/OneDrive/MScthesis/data/comm.csv",",",)%>%
   filter(id!="Standard 2")
 
 
 qpcr <- qpcr %>% 
   inner_join(std, by = c("time" = "time", "type" = "type"))%>%
   mutate(ct_corr = ct + corr)%>%
   select(time,type,id,ct_corr)%>%
   mutate(qty= 10^((ct_corr-40.36)/-3.648))
  
 rm(std)
 rm(standard_curve)
 rm(model)
```



# Basic statistics and data mangling   
```{r sum stats and normalization , echo=FALSE, warning=FALSE, message = FALSE}
# Summary statistics

 summary <- qpcr %>%
    group_by(type, id)%>%                                    
    summarise(
      n=n(),
      avg.qty = mean(qty, na.rm=TRUE),
      sd.qty=sd(qty, na.rm = TRUE) 
    )%>% 
    ungroup()
 summary
 rm(qpcr)
 
# Join DNA conc. and mean genecopy quantity for later calculations (to get genecopy number per ng DNA)  
 
 joined <- summary %>% 
   inner_join(dna, by = c("id" = "id", "type" = "type"))%>% 
   rename(dna_conc= value)
 rm(summary)
 rm(dna)
 
#Normalize quantity values
  
  # Reset dilution
  # Normalize on 1 ul (8 ul template used in one well)
  # Normalize on 1 ng DNA (divide by DNA conc. per ng)
  # Index created: averaged quantity value (from technical replicates) for each replicate
 
   calc <- joined %>%
     mutate(qty_undil=avg.qty*400)%>% 
     mutate(qty_ul=qty_undil/8)%>% 
     mutate(qty_ng=qty_ul/dna_conc)
   calc
   
   index <- calc  %>% 
    inner_join(trt, by = c("id" = "id"))%>%
    arrange(type, treatment)
    
   
   rm(joined)
   
  # From calc we extract the calculated values into a new tibble and remove calc 
    # There are other columns in calc which contain unwanted information, hence the new tibble
    # You could also set the unwanted columns to NULL and work further with calc 
   
  fin <- calc %>% 
     select(type,id, qty_ng) %>% 
     mutate(unit = "GeneCopies/ngDNA")
  rm(calc)
  
  # Add descriptors (another table) for treatment factors
  
  genecopy <- fin %>% 
   inner_join(trt, by = c("id" = "id"))
  genecopy
  
  rm(trt)
  rm(fin)
  
  
# Finished, we now have a tibble called 'genecopy' which includes the gene copy numbers per ng DNA
  # Index is a copy of all average qty values (avg of technical replicates) for each replicate and is normalized on ng DNA


```

The data preparation is finished for the commamox gene copy data. We have worked with the raw quantity values used qPCR machine (standardized on the individual runs standard curve), did summary statistics (averaging the technical replicates) and calculated gene copy number per nanogram DNA. 

Now we move on to the statistical modelling and checking for significance.

# Statistical modelling and sig. testing {.tabset .tabset-fade .tabset-pills}

## Rhizo

### Checking data balance

```{r rhizo commamox group means and see if data is balanced, echo=FALSE}
# Needs the genecopy file from the 16S data mangling qPCR chunk 
# Following the instructions of        
  # https://www.datanovia.com/en/lessons/repeated-measures-anova-in-r/
  # http://coleoguy.github.io/teaching/expdes/lec8.pdf
  # https://www.youtube.com/watch?v=i7tJwdZrklM&feature=youtu.be
  
# Start with analysis of rhizo
  
genecopy.work <- genecopy %>%
  filter(type =="rhizo")%>%
  ungroup()

# Deleting columns that are not needed
genecopy.work$unit <- NULL
genecopy.work$treatment <- NULL
genecopy.work$type <- NULL

#Inference about dataset, see if ANOVA assumptions are met

  #Summary statistics a to get a feel of our data 
  genecopy.work %>%
  group_by(fert, fung, time) %>%
  get_summary_stats(qty_ng, type = "mean_sd")
```

We have a balanced dataset (n is the same for each group). Moving on to a boxplot to visually inspect groups and look for outliers.

```{r rhizo commamox boxplots, echo=FALSE}
  bxp <- ggboxplot(
    genecopy.work, x = "fert", y = "qty_ng",
    color = "time", palette = "jco",
    facet.by = "fung", short.panel.labs = FALSE
  )
  bxp+ggtitle("Rhizo: commamox gene copy numbers")
```

We have each one outlier in '[fert:no|fung:no|time:t3]', '[fert:yes|fung:no|time:t2]' and '[fert:yes|fung:yes|time:t3]'. Let's check if they are extreme:

```{r rhizo commamox outliers, echo=FALSE} 
  outlier <-  genecopy.work %>%
    group_by(fert,fung, time) %>%
    identify_outliers(qty_ng)
  outlier
```

According to the definition

> Values above Q3 + 1.5xIQR or below Q1 - 1.5xIQR are considered as outliers. Values above Q3 + 3xIQR or below Q1 - 3xIQR are considered as extreme points (or extreme outliers).

there are no extreme outliers in this dataset. 

###### Datatable from group [fert:yes|fung:yes|time:t3]
```{r [fert:no|fung:no|time:t2], echo = FALSE}
inspect<- genecopy.work %>%
  filter(fert =="yes" & fung == "yes" & time == "t3")
inspect
rm(inspect)
``` 

### Normality and heteroskedasticity

```{r rhizo 16s normality test and qq plots, echo=TRUE}
  # Normality test with all outliers (Shapiro Test)
    
  genecopy.work %>%
    filter (id!="330")%>%
    group_by(fert, fung, time) %>%
    shapiro_test(qty_ng)

# QQplots: Investigate heteroskedasticity visually 

  # All datapoints combined
   genecopy.work %>%
    ggqqplot("qty_ng", ggtheme = theme_bw())
   
  # Grouped by treatments and timepoints
   genecopy.work %>%
    ggqqplot("qty_ng", ggtheme = theme_bw())+
    facet_grid(fert + fung ~ time, labeller = "label_both")
   
# The deviation in [fert:no|fung:no|time:t3] is clearly visible, exclude the outlier [id: 355]
  # [fert:yes|fung:yes|time:t2] shows also an outlier far away from the line, somewhere at 3e+05 [id:253]
  # Repeat normality test with filter of outliers
  # Heteroskedascticity not perfect but OK 
   
  genecopy.work.filtered <- genecopy.work %>%
    filter (id!="330")
```

All p-values above 0.05. We can proceed without filtering any replicates. 330 excluded

### ANOVA: choosing a  model 

An analysis of the dataset structure is needed to find the right statistical model. The data was generated in a randomized complete block design (RCBD). 

##### Model parameters {.tabset .tabset-fade .tabset-pills}

###### Fixed effects 

* Fertilizer (fert) (Binary variable (yes/no))
* Fungicide and growth (fung) (Binary variable (yes/no))

###### Random effects

* Time 
* Block
* (Row)
* (Could possibly include rainfall 3-7 days before sampling)


#### Mixed effect model

The model needs to account for the above listed fixed and random effects. The regular lm() function of R stats package will fit all variables as fixed effects if they are integrated into the forumlae. Therefore we need the package nlme which can account for random effects. Because we are using only two timepoints I will stick to a linear model. Generally, I'd consider a non-linear model if all timepoints would be in the analysis. We are investigating gene copy numbers which are directly correlated and have a causal relation ship with number of bacteria. Bacteria growth is better estimated with a logistic regression. 

I'll do a stepwise modelling approach without the mathematic forumlae (will be done in the thesis tho).

##### All main effects + Interaction (Full model)

```{r rhizo commamox: mixed effect model  }
# Fitting a linear mixed effect model to our genecopynumber dataset
  # Treatment variables (fert|fung) are treated as fixed effects
  # Time and Block are random effect variables (can't be replicated)
  # We use the package nlme to effectively input random effect variables for our model
  # Interaction of fert*fung must be tested, because we have two factorial experiment 
  # It is also possible to just type in fert*fung as predictor variable, the package will fit also the main effects
  # If more timepoints of this are used, a non-linear model would be better because bacteria growth is not linear
  
  fit.all <- lme(qty_ng ~ fert+fung+fert*fung,
                  random = list(~1|block, ~1|time),
                  data=genecopy.work.filtered)
  summary(fit.all)
  

```

The interaction term is not significant, so I'm droping it from the model, leaving only the main effects in.

##### Both main effects (Reduced model)

```{r rhizo 16S mixed effect model no interaction}  
  fit.nointeraction <- lme(qty_ng ~ fert+fung,
                  random = list(~1|block, ~1|time),
                  data=genecopy.work.filtered)
  summary(fit.nointeraction)
```

Fungicide is highly sig. while fertilizer is not sig. Dropping fert from the model.


##### Only Fung as main effect

```{r rhizo 16S mixed effect model no interaction no fung} 
  
  fit.nofert <- lme(qty_ng ~ fung,
                  random = list(~1|block,~1|time),
                  data=genecopy.work.filtered)
  summary(fit.nofert)
```
##### Only Fert as main effect

```{r rhizo 16S mixed effect model no interaction no fung} 
  
  fit.nofung <- lme(qty_ng ~ fert,
                  random = list(~1|block, ~1|time),
                  data=genecopy.work.filtered)
  summary(fit.nofung)
  
```

## Soil 

### Checking the datas balance 
```{r soil 16s group means and see if data is balanced, echo=FALSE}
# Needs the genecopy file from the 16S data mangling qPCR chunk 
# Following the instructions of        
  # https://www.datanovia.com/en/lessons/repeated-measures-anova-in-r/
  # http://coleoguy.github.io/teaching/expdes/lec8.pdf
  # https://www.youtube.com/watch?v=i7tJwdZrklM&feature=youtu.be
  
genecopy.work <- genecopy %>%
  filter(type =="soil")%>%
  ungroup()

# Deleting columns that are not needed
genecopy.work$unit <- NULL
genecopy.work$treatment <- NULL
genecopy.work$type <- NULL

#Inference about dataset, see if ANOVA assumptions are met

  #Summary statistics and boxplots, to get a feel of our data and check visually for outliers
  genecopy.work %>%
  group_by(fert, fung, time) %>%
  get_summary_stats(qty_ng, type = "mean_sd")
```

We have a balanced dataset (n is the same for each group). Moving on to a boxplot to visually inspect groups and look for outliers.

### Boxplot to inspect data visually

```{r soil 16s boxplots, echo=FALSE}
  bxp <- ggboxplot(
    genecopy.work, x = "fert", y = "qty_ng",
    color = "time", palette = "jco",
    facet.by = "fung", short.panel.labs = FALSE
  )
  bxp + ggtitle("Soil: commamox gene copy numbers ")
```  

We have one outlier in '[fert:no|fung:no|time:t2]', and two in '[fert:yes|fung:yes|time:t3]'. The boxplot of [fert:no|fung:yes|time:t2] is very broad, which the SD of the previous table already predicted. Lets check the outliers.
### Identifying the outliers.

```{r soil commamox outliers, echo=TRUE}    
  outlier <-  genecopy.work %>%
    group_by(fert,fung, time) %>%
    identify_outliers(qty_ng)
  outlier
```

One outlier in '[fert:yes|fung:yes|time:t3]' is extreme. 

> Values above Q3 + 1.5xIQR or below Q1 - 1.5xIQR are considered as outliers. Values above Q3 + 3xIQR or below Q1 - 3xIQR are considered as extreme points (or extreme outliers). 

Lets check the groups datatable:

##### Groups that failed the Shapiro Test and their replicates  {.tabset}
###### Datatable from group [fert:yes|fung:yes|time:t3]
```{r [fert:no|fung:no|time:t2], echo = FALSE}
inspect<- genecopy.work %>%
  filter(fert =="yes" & fung == "yes" & time == "t3")
inspect
rm(inspect)
``` 


##### {-}
### Normality and heteroskedasticity

```{r soil commamox normality test and qq plots, echo=FALSE}
# Normality test with all outliers (Shapiro Test)
  genecopy.work %>%
    group_by(fert, fung, time) %>%
    shapiro_test(qty_ng)

# QQplots: Investigate heteroskedasticity visually 

  # All datapoints combined
   genecopy.work %>%
    ggqqplot("qty_ng", ggtheme = theme_bw())
   
  # Grouped by treatments and timepoints
   genecopy.work %>%
    ggqqplot("qty_ng", ggtheme = theme_bw())+
    facet_grid(fert + fung ~ time, labeller = "label_both")

```

All groups pass the normality test. QQplots look reasonably heteroskedastic. Has a bit of a tail, more extreme values. 


### ANOVA: Choosing a model

An analysis of the dataset structure is needed to find the right statistical model. The data was generated in a randomized complete block design (RCBD). 

##### Model parameters {.tabset .tabset-fade .tabset-pills}

###### Fixed effects 

* Fertilizer (fert) (Binary variable (yes/no))
* Fungicide and growth (fung) (Binary variable (yes/no))

###### Random effects

* Time 
* Block
* (Row)
* (Could possibly include rainfall 3-7 days before sampling)


#### Mixed effect model

The model needs to account for the above listed fixed and random effects. The regular lm() function of R stats package will fit all variables as fixed effects if they are integrated into the forumlae. Therefore we need the package nlme which can account for random effects. Because we are using only two timepoints I will stick to a linear model. Generally, I'd consider a non-linear model if all timepoints would be in the analysis. We are investigating gene copy numbers which are directly correlated and have a causal relation ship with number of bacteria. Bacteria growth is better estimated with a logistic regression. 

I'll do a stepwise modelling approach without the mathematic forumlae (will be done in the thesis tho).

##### All main effects + Interaction (Full model)

```{r soil commamox anova full model, echo = FALSE}
# Fitting a linear mixed effect model to our genecopynumber dataset
  # Treatment variables (fert|fung) are treated as fixed effects
  # Time and Block are random effect variables (can't be replicated)
  # We use the package nlme to effectively input random effect variables for our model
  # Interaction of fert*fung must be tested, because we have two factorial experiment 
  # It is also possible to just type in fert*fung as influencer variable, the package will fit also the main effects
  # If more timepoints of this are used, a non-linear model would be better because bacteria growth is not linear
  
  fit.all <- lme(qty_ng ~ fert+fung+fert*fung,
                  random = list(~1|block, ~1|time, ~1|row),
                  data=genecopy.work)
  summary(fit.all)
  
```
Interaction is not sig. Dropping it from the model
##### Reduced model (both main effects)

```{r soil commamox anova reduced model(both main effects), echo = FALSE}
  # The interaction term is not significant, so drop it from the model 
  
  fit.nointeraction <- lme(qty_ng ~ fert+fung,
                  random = list(~1|block, ~1|time),
                  data=genecopy.work)
  summary(fit.nointeraction)
```
Fertilizer is highly significant. Fungicide is borderline significant. May be worth investigating.

##### Reduced model 2 (only fert)

```{r soil 16S anova reduced model(only fert), echo = FALSE}
  
  fit.nofung <- lme(qty_ng ~ fert,
                  random = list(~1|block,~1|time),
                  data=genecopy.work)
  summary(fit.nofung)
```
```{r}

genecopy$subject <- as.factor(genecopy$subject)

fit <- lmer(qty_ng ~  
              subject + (1|block),
            data=genecopy)
summary(fit)

```
```{r}
fm1 <- lmer(qty_ng ~ fert*fung*time  + (1|block), data=genecopy.work)
```



