---
title: "MaTh - Material and Methods"
author: "Joshua Johnson"
output: html_notebook
---
```{r Packages, echo=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(cowplot)
library(reshape2)
```

# Climate data 

Close to experimental Site (53°21'58.5"N|13°48'13.3"E). Data extracted from wetterkontor.com from weather station in Grünow (53°19'01.7"N|13°56'55.0"E)
```{r Climate Data}

#Read data as tibble (tidyverse)
  climate <- 
    read_delim("C:/Users/jjohn/OneDrive/MScthesis/data/clim.csv",";",) %>%
    filter (date != 0)#deletes empty days (artifact from data mining)
  
  climate$date <- as.Date(climate$date,"%d.%m.%Y") #formats the date correct 

#Summary statistics as we have day data, condense to month 
  summary <- climate %>%
    mutate(month = format(date,"%m"), year = format (date, "%y"))%>% #new month and year var
    group_by(id, month, year)%>%                                    #group by new vars
    summarise(
      avg = mean(value) #mean it 
    )

# Plotting the weather data 

  summary$id <- 
    as.factor(summary$id)
  summary$time <- 
    lubridate::ymd(paste0(summary$year,summary$month,"01"))#reintroducing date format for ggplot2

# consistent coloring scheme
  my_color <- c ("deepskyblue1", "goldenrod1", "black", "red", "dodgerblue4") 
  names(my_color) <- levels(summary$id)
  my_scale <- scale_color_manual(name = "Legend", 
                                 values = my_color,
                                 breaks=c("temp_max","temp_avg", "temp_min"),
                                 labels=c("Maximum", "Average", "Minimum"))  
  
# filtering data for separate plots
  
  temp <- filter(summary, id =="temp_max" |id =="temp_min" | id=="temp_avg") 
  sun <- filter(summary,id =="sunshine")
  rain<- filter(summary,id=="rainfall")
  

# ggplot area
  #temperature
  temp$title <- "Temperature"
  a <- ggplot(temp, aes(time,avg,color=id)) +
    geom_line() +
    ylab("°C")+
    xlab("2019-2020")+
    facet_grid(~title)+
    NULL 
  plot_temp <- a + my_scale
  
  #rainfall, sunshine
  rain$title <- "Rainfall" 
  b <- ggplot(rain, aes(time,avg, color=id)) +
    geom_line() +
    ylab(expression(paste("L/m"^"2")))+
    xlab("2019-2020")+
    facet_grid(~title)+
    NULL 
  plot_rain <- b + my_scale

  sun$title <- "Sunshine"
  c <- ggplot(sun, aes(time,avg,color=id)) +
    geom_line() +
    ylab("Hours")+
    xlab("2019-2020")+
    facet_grid(~title)+
    NULL
  plot_sun <- c + my_scale
    

#cowplot to arrange 
  #Plotting two plots together
  plot_other<- plot_grid(plot_rain + theme(legend.position="none"),
                         plot_sun + theme(legend.position="none")
                         ,labels = c('B', 'C'))
  # so they can be in one row 
  prow <- plot_grid(plot_temp,
            plot_other,
            labels = c('A', ''),
            ncol = 1, nrow = 2)
  prow
  
  legend <- get_legend(
 


```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
