---
title: "ChemLy"
author: "Joshua Johnson"
date: "1/8/2021"
output: 
  pdf_document:
    extra_dependencies: ["mhchem"]
header-includes:
  - \usepackage[version=4]{mhchem}
---

```{r setup, include=FALSE, echo=FALSE, results='hide', warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(ggpubr)
```

```{r data / labs load, echo=FALSE}

# Read data as tibble (tidyverse)

  chem <- 
      read_delim("C:/Users/jjohn/OneDrive/MScthesis/data/chemanalysis.csv",",",) %>%
      filter (crop == "W") #only need W for my thesis

# Correct Variable types here 

  chem$value <- as.numeric(chem$value)
  chem$id    <- as.factor(chem$id)
  chem$treatment <- as.factor(chem$treatment)

#Summary statistics 

  summary <- chem %>%
    group_by(type, timepoint, id, treatment)%>%                     
    summarise(
      n = n(),
      avg = mean(value),
      sd = sd(value),
      med = median(value),
      min = min(value),
      max = max(value),
    )

  

  
  # General facet labels for better understanding (Abbreviation index)
  
   labs <- c(
     c_t= "Total Carbon",
     n_t="Total Nitrogen",
     nit = "Nitrate",
     amo = "Ammonium",
     ph = "pH",
     H00 = "Control",
     H01 = "Fertilizer",
     HP0 = "Fung+GroReg",
     HP1 = "Full",
     t2 = "t1",
     t3 = "t2"
     
            )
   
```


## Chemical analysis of soil and roots 

Nitrogen was extracted with \ce{KCl} from wet soil samples from first and second timepoint. Available nitrate (\ce{NO3-}) and ammonium (\ce{NH4+}) were measured measured by ion chromatography. 

Plant-available nitrogen in the forms of nitrate (\ce{NO3-}) and ammonium (\ce{NH4+}) were measured in wet soil samples. The data is visualized in barplots with the  standard deviation 

```{r soil nitrogen, echo=FALSE}

  # Plant-available nitrogen
  #Nitrate
  reduced <- filter(summary,id == "nit") #filter so we got what we need
  
  plot.nit <- ggplot(reduced,aes(timepoint,avg,fill=timepoint))+
    geom_bar(stat="identity", 
             position=position_dodge()
             )+
    geom_errorbar(aes(ymin=avg-sd, ymax=avg+sd), width=.2,
                 position=position_dodge(.9))+
    ylab("mg in 100g wet soil")+
    facet_grid(treatment~id,
               scales="free",
               labeller = labeller(id = labs, treatment = labs, timepoint = labs)
               )
   
    NULL
  #Ammonium
  reduced <- filter(summary, id == "amo" )
  plot.amo <- ggplot(reduced,aes(timepoint,avg,fill=timepoint))+
    geom_bar(stat="identity", 
             position=position_dodge()
             )+
    geom_errorbar(aes(ymin=avg-sd, ymax=avg+sd), width=.2,
                 position=position_dodge(.9))+
    ylab("mg in 100g wet soil")+
    facet_grid(treatment~id,
               scales="free",
               labeller = labeller(id = labs, treatment = labs, timepoint = labs)
               )
   
    NULL
  plot.nitamo <- ggarrange(plot.nit, plot.amo, legend = FALSE)
```



```{r Chemical Analysis, echo=FALSE, results='hide'}

  
  
  # C_t, n_t soil
  
  reduced <- filter(summary, type =="soil", id == "c_t" | id == "n_t")
  
  plot.ctnt <- ggplot(reduced,aes(treatment,avg,fill=treatment))+
    geom_bar(stat="identity", 
             position=position_dodge()
             )+
    geom_errorbar(aes(ymin=avg-sd, ymax=avg+sd), width=.2,
                 position=position_dodge(.9))+
    NULL
  
  plot.ctnt + 
    facet_wrap(~id,
               scales="free",
               labeller = labeller(id = labs)
               ) +
    ylab("[% of dried soil sample]")+
    labs("Concentration of carbon and nitrogen in dried soil")
  
  # C_t, n_t root
  
  reduced <- filter(summary, type =="root", id == "c_t" | id == "n_t")
  
  plot.ctnt.rt <- ggplot(reduced,aes(treatment,avg,fill=treatment))+
    geom_bar(stat="identity", 
             position=position_dodge()
             )+
    geom_errorbar(aes(ymin=avg-sd, ymax=avg+sd), width=.2,
                 position=position_dodge(.9))+
    NULL
  
  plot.ctnt.rt + 
    facet_wrap(timepoint~id,
               scales="free",
               labeller = labeller(id = labs)
               ) +
    ylab("[% of dried root sample]")+
    labs(title = "Total carbon and nitrogen content in dried roots")
  
  # pH
  # I added a title here to get the facet_wrap title 
  
  reduced <- filter(summary, id == "ph")
  reduced$title <- "pH"
  
  plot.ph <- ggplot(reduced,aes(treatment,avg,fill=treatment))+
    geom_bar(stat="identity", 
             position=position_dodge()
             )+
    geom_errorbar(aes(ymin=avg-sd, ymax=avg+sd), width=.2,
                 position=position_dodge(.9))+
    facet_wrap(~title)+
    NULL
  
  plot.ph+
    ylab(expression("-log [H"^"+"*"]"))+
    labs(title = "pH in Soil")
  
```